\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{fancyvrb}

\title{CS257 Distributed Database\\Milestone 1 Status Report}
\author{Team: Distributed Database Project}
\date{\today}

\begin{document}
\maketitle

\section{Team Organization}
\begin{itemize}[leftmargin=1.2em]
    \item \textbf{Gautam Panigrahi} --- primary developer; responsible for architecture design, coordinator/shard service implementation, Docker-based deployment, and initial unit testing.
    \item \textbf{Current operating model} --- single-developer execution for Sprint 1 prototype.
    \item \textbf{Planned scaling} --- if additional teammates are added, roles will be split into (i) transaction/consensus, (ii) fault tolerance/recovery, and (iii) test/benchmark automation.
\end{itemize}

\section{Server/System Software/Hardware Configuration}
\subsection*{Software Stack}
\begin{itemize}[leftmargin=1.2em]
    \item Language/runtime: Python services (containerized with \texttt{python:3.11-slim}).
    \item API framework: FastAPI.
    \item HTTP client for inter-node RPC: \texttt{httpx}.
    \item Validation/models: Pydantic.
    \item Service host for APIs: Uvicorn.
    \item Container orchestration: Docker Compose.
    \item Testing: Pytest + FastAPI TestClient.
\end{itemize}

\subsection*{Cluster Configuration (Prototype)}
\begin{itemize}[leftmargin=1.2em]
    \item 5 services total: 1 coordinator + 2 leaders + 2 followers.
    \item Coordinator exposed at \texttt{localhost:8000}.
    \item Shard leaders exposed at \texttt{localhost:8001} and \texttt{localhost:8003}.
    \item Shard followers exposed at \texttt{localhost:8002} and \texttt{localhost:8004}.
    \item Sharding function: \texttt{shard\_id = sha256(key) \% N} with \texttt{N=2} for Sprint 1.
\end{itemize}

\subsection*{Hardware/Host Notes}
\begin{itemize}[leftmargin=1.2em]
    \item Development host environment: macOS.
    \item Local validation in this environment confirmed repository cloning and source inspection.
    \item Team should add final machine specs (CPU/RAM/storage) used for benchmark runs in Milestone 2.
\end{itemize}

\section{Project Definition}
\subsection*{Goal}
Build a minimal distributed key-value database that demonstrates key distributed systems primitives: deterministic sharding, coordinator-based routing, replication, and transactional behavior.

\subsection*{Functional Requirements}
\begin{itemize}[leftmargin=1.2em]
    \item Route client requests through a single coordinator API.
    \item Deterministically map keys to shards using stable hashing.
    \item Support transaction lifecycle endpoints: \texttt{/begin}, \texttt{/write}, \texttt{/read}, \texttt{/commit}, \texttt{/abort}.
    \item Provide read-your-own-writes behavior within a transaction.
    \item Replicate committed updates from leader to follower.
    \item Expose node/cluster observability endpoints (e.g., \texttt{/cluster}, \texttt{/health}, \texttt{/data}).
\end{itemize}

\subsection*{Non-Functional Requirements}
\begin{itemize}[leftmargin=1.2em]
    \item \textbf{Consistency}: commit should fail if follower replication does not acknowledge.
    \item \textbf{Determinism}: all nodes must compute identical shard mapping for a key.
    \item \textbf{Modularity}: clear separation of coordinator, shard node logic, and shared utilities.
    \item \textbf{Reproducibility}: one-command startup via Docker Compose.
    \item \textbf{Testability}: include automated tests for hashing and shard transaction behavior.
\end{itemize}

\section{Architecture}
\subsection*{Architecture Outline}
\begin{itemize}[leftmargin=1.2em]
    \item Clients interact only with the coordinator service.
    \item Coordinator computes target shard for key-based operations and forwards RPC calls to the corresponding leader.
    \item Each shard has one leader and one follower.
    \item On commit, leader first replicates updates to follower, then applies updates locally.
    \item Coordinator tracks shards touched by each transaction and issues per-shard commit/abort requests.
\end{itemize}

\subsection*{Architectural Diagram}
\begin{Verbatim}[fontsize=\small,frame=single]
             +----------------+
 clients --> |  Coordinator   |  :8000
             +-------+--------+
                     |
      route: shard_id = sha256(key) % N
             +-------+--------+
             |                |
      +------v------+   +-----v-------+
      | shard0      |   | shard1      |
      | leader:8001 |   | leader:8003 |
      +------+------+
             | sync replicate on commit
      +------v------+   +-----v-------+
      | follower    |   | follower    |
      |    :8002    |   |    :8004    |
      +-------------+   +-------------+
\end{Verbatim}

\section{Scope So Far}
\subsection*{Research/Evaluation Methodologies}
\begin{itemize}[leftmargin=1.2em]
    \item Incremental prototype approach: implement narrow, end-to-end vertical slices first (routing + single-shard transactions + replication).
    \item Unit-test-first validation for critical primitives: hash stability/distribution and shard transaction semantics.
    \item API smoke testing methodology documented via \texttt{curl} and \texttt{demo.py} walkthrough.
    \item Containerized deployment used to reduce environment drift and support repeatable team demos.
\end{itemize}

\subsection*{Implementation Details}
\begin{itemize}[leftmargin=1.2em]
    \item \texttt{common/hashing.py}: stable \texttt{sha256}-based shard function to avoid Python hash randomization.
    \item \texttt{common/config.py}: environment-driven shard topology resolution.
    \item \texttt{coordinator/main.py}: transaction registry, request routing, cluster metadata endpoint.
    \item \texttt{shard/main.py}: in-memory committed store + per-transaction staging area, role-based leader/follower behavior.
    \item \texttt{docker-compose.yml}: bootstraps full 5-node prototype with explicit role/env config.
\end{itemize}

\subsection*{Requirements Implemented Thus Far}
\begin{itemize}[leftmargin=1.2em]
    \item Deterministic sharding and coordinator routing.
    \item Transaction begin/write/read/commit/abort APIs.
    \item Read-your-own-writes semantics before commit.
    \item Leader-to-follower replication path on commit.
    \item Commit guardrail: leader does not apply updates if follower replication fails.
    \item Unit test coverage for key hashing and shard transactional behavior.
\end{itemize}

\subsection*{Preliminary Results}
\begin{itemize}[leftmargin=1.2em]
    \item Test suite was executed in a local Python 3.11 virtual environment (\texttt{.venv311}) with result: \texttt{8 passed in 0.41s}.
    \item Live coordinator API checks were executed against a running Docker Compose cluster at \texttt{localhost:8000}; captured outputs are shown in the prototype snapshots below.
    \item Current design provides strong leader/follower consistency within a shard commit path, but cross-shard commits remain best-effort and can be partial (known Sprint 1 limitation).
\end{itemize}

\section{Prototype (Snapshots of a Running Prototype)}
\textbf{Snapshot A: Cluster topology endpoint (captured from running stack)}
\begin{Verbatim}[fontsize=\small,frame=single]
GET http://localhost:8000/cluster

{"num_shards":2,"shards":{"0":{"leader":"http://shard0-leader:8000",
"follower":"http://shard0-follower:8000"},"1":{"leader":"http://shard1-leader:8000",
"follower":"http://shard1-follower:8000"}},"hash_scheme":"sha256(key) % num_shards",
"active_txns":0}
\end{Verbatim}

\textbf{Snapshot B: Transaction write/read/commit flow (captured)}
\begin{Verbatim}[fontsize=\small,frame=single]
txn_id=c9fba047-87e5-42b8-a077-1672228dc793

POST /write:
{"ok":true,"shard_id":0,"result":{"ok":true,"node":"shard0-leader",
"staged":{"user:1":"alice"}}}

POST /read (in-txn):
{"key":"user:1","value":"alice","source":"staged",
"node":"shard0-leader","shard_id":0}

POST /commit:
{"ok":true,"committed_shards":[0],"results":{"0":{"ok":true,
"applied":1,"node":"shard0-leader"}}}

POST /read (post-commit):
{"key":"user:1","value":"alice","source":"committed",
"node":"shard0-leader","shard_id":0}
\end{Verbatim}

\textbf{Snapshot C: Abort behavior (captured)}
\begin{Verbatim}[fontsize=\small,frame=single]
txn2_id=019bbebd-ab2c-41f7-80d1-96c64106b66b

POST /write:
{"ok":true,"shard_id":0,"result":{"ok":true,"node":"shard0-leader",
"staged":{"temp:1":"X"}}}

POST /abort:
{"ok":true,"aborted_shards":[0]}

POST /read (after abort):
{"key":"temp:1","value":null,"source":"missing",
"node":"shard0-leader","shard_id":0}
\end{Verbatim}

\section{Next Steps}
\begin{enumerate}[leftmargin=1.4em]
    \item Implement two-phase commit (2PC) for atomic multi-shard transactions.
    \item Add strict concurrency control (e.g., lock manager / strict 2PL or MVCC-based variant).
    \item Add write-ahead logging and recovery procedures for crash tolerance.
    \item Implement leader failure handling and follower promotion.
    \item Build benchmark harness for throughput/latency under different shard counts and contention levels.
    \item Expand test suite to include coordinator integration tests and fault-injection scenarios.
\end{enumerate}

\section{Risks and Mitigation Plans}
\begin{itemize}[leftmargin=1.2em]
    \item \textbf{Risk: Partial cross-shard commit failures} (no 2PC yet).\\
    \textbf{Mitigation:} prioritize 2PC with explicit prepare/commit phases and rollback strategy.

    \item \textbf{Risk: Data loss on crash} (in-memory state only).\\
    \textbf{Mitigation:} add durable WAL, checkpointing, and restart recovery flow.

    \item \textbf{Risk: Single-point coordinator bottleneck/failure}.\\
    \textbf{Mitigation:} coordinator replication or stateless coordinator with external metadata/log service.

    \item \textbf{Risk: Replication coupling reduces availability}.\\
    \textbf{Mitigation:} make consistency mode configurable (strict sync vs quorum/async for future experiments).

    \item \textbf{Risk: Limited observability for debugging distributed behavior}.\\
    \textbf{Mitigation:} add structured logs, trace IDs per transaction, and lightweight metrics dashboard.
\end{itemize}

\section*{Appendix: Repository and Run Commands}
Repository: \url{https://github.com/gautampanigrahi/distributed-database}

\begin{Verbatim}[fontsize=\small,frame=single]
docker compose up --build
python demo.py
\end{Verbatim}

\end{document}
